From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-04 Pi+/paper/finite.tex
Cut more in section 3

However, more conveniently, in HoTT, we can instead use groupoids to talk about groups internally. A group can be
identified with a 1-object groupoid, using a technique called delooping. The delooping of a group $G$ is a groupoid
$\B{G}$, with a unique object $\pt$, where the group elements are encoded as self-loops $\pt \id_{\B{G}} \pt$.

The group operations are automatically given by operations on the identity type, with $\refl_{\pt}$ for the neutral
element, path composition for the group multiplication, and path inverse for the group inverse. These satisfy the group
laws, up to the identity type, using the groupoid coherence laws.

Since groups are sets, these 2-paths are propositions, making $\B{G}$ a 1-groupoid. Hence, a group is simply given by a
pointed, connected 1-type~\cite*{buchholtzHigherGroupsHomotopy2018,symmetryBook2021}. \footnote{This technique also
  allows defining higher groups in HoTT, as done in~\cite{buchholtzHigherGroupsHomotopy2018}.}

>>>>

% However, more conveniently, in HoTT, we can instead use groupoids to talk about groups internally. A group can be
% identified with a 1-object groupoid, using a technique called delooping. The delooping of a group $G$ is a groupoid
% $\B{G}$, with a unique object $\pt$, where the group elements are encoded as self-loops $\pt \id_{\B{G}} \pt$.

% The group operations are automatically given by operations on the identity type, with $\refl_{\pt}$ for the neutral
% element, path composition for the group multiplication, and path inverse for the group inverse. These satisfy the group
% laws, up to the identity type, using the groupoid coherence laws.

% Since groups are sets, these 2-paths are propositions, making $\B{G}$ a 1-groupoid. Hence, a group is simply given by a
% pointed, connected 1-type~\cite*{buchholtzHigherGroupsHomotopy2018,symmetryBook2021}. \footnote{This technique also
%   allows defining higher groups in HoTT, as done in~\cite{buchholtzHigherGroupsHomotopy2018}.}

### 2021-07-04 Pi+/paper/ufin.tex
Cut more in section 3

HoTT provides a rich internal language for working with weak higher groupoids, using the ``Types are weak
$\infty$-groupoids'' correspondence. Using the identity type for morphisms, a 1-groupoid in HoTT has points for 0-cells,
1-paths for 1-cells, and 2-paths for 2-cells, but there is at most one 2-cell between compatible 1-cells.~\footnote{This
  is a locally-strict $(2,0)$-category, since every cell above 0 is invertible, and the hom-categories are posets, that
  is, truth-value enriched.} The groupoid $\BFin$ can be written in HoTT as a type $\UFin$, which is a 1-groupoid.

@@

\subsubsection{Identity Types}

@@

\begin{toappendix}
  \begin{definition}[Path Induction]
    \label{def:path-induction}
    Given a type family $C : \dfun{x,y:A}{(x \id_A y)} \to \UU$, and a function $c : \dfun{x:A}{C(x,x,\refl_x)}$, there is
    a function $f : \dfun{x,y:A}{\dfun{p:x \id_A y}{C(x,y,p)}}$ such that $f(x,x,\refl_x) \defeq c(x)$.
  \end{definition}
\end{toappendix}

@@

Further, using the groupoid structure of $A$, we can show that $\transport{P}{p}{\blank}$ and
$\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name the map $\tptEqv{P}$ which lifts paths to equivalences
of fibers.

@@

\todo{Diagram.}
The type families (or fibrations) we're interested in are the ones where paths in the base space completely determine
the equivalences in the fibers -- these are called univalent fibrations.

@@

\begin{toappendix}
  \begin{definition}[Propositional Truncation]
    \label{def:prop-trunc}
    Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, or simply $\Trunc{A}$, is a higher inductive type
    generated by the following constructors,
    \begin{itemize}
      \item an inclusion function $\trunc{\blank} : A \to \Trunc{A}$,
      \item for each $x, y : \Trunc{A}$, a path $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$,
    \end{itemize}
    such that, given any type $B$ with
    \begin{itemize}
      \item a function $g : A \to B$,
      \item for each $x, y : B$, a path $\term{trunc*}(x,y) : x \id_{B} y$,
    \end{itemize}
    there is a unique function $f : \Trunc{A} \to B$ such that,
    \begin{itemize}
      \item $f(\trunc{a}) \equiv g(a)$
      \item for each $x, y : \Trunc{A}$, $\ap{f}{\term{trunc}(x,y)} \id_{B} \term{trunc*}(f(x),f(y))$.
    \end{itemize}
  \end{definition}
\end{toappendix}

@@

\begin{toappendix}
  \begin{definition}[Set Quotient]
    \label{def:set-quot}
    Given a type $A$ which is an $\hSet$, and a relation $R : A \to A \to \hProp$, the set-quotient $\quot{A}{R}$ is the
    higher inductive type generated by
    \begin{itemize}
      \item an inclusion function $q : A \to \quot{A}{R}$,
      \item for each $x, y : A$ such that $R(x,y)$, a path $q(x) \id_{\quot{A}{R}} q(y)$,
      \item a set truncation, for each $x, y : \quot{A}{R}$ and $r, s : x \id_{\quot{A}{R}} y$, we have $r \id s$,
    \end{itemize}
    with \review{an appropriate induction principle.}
  \end{definition}
\end{toappendix}

@@

Starting from a univalent universe which classifies all types, we want to define a subuniverse which classifies only
certain types, for example, types that satisfy some desired property. We use a prop-valued type family, that is, a
predicate, which picks out only those types, and collect them into a univalent subuniverse. Being univalent ensures that
the equality type of the ambient universe is reflected in the subuniverse.

@@

\begin{definition}[Subuniverse]
  A subtype is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $P(X)$ is a proposition for every
  $X$. A subuniverse generated by a subtype has $U \defeq \dsum{X:\UU}{P(X)}$ and $\El \defeq \fst$.
\end{definition}

\begin{proposition}[Univalent Subuniverse]
  Subuniverses generated by subtypes are univalent.
\end{proposition}

@@

The types we're interested in are the finite types. In constructive mathematics, the notion of finiteness is subtle. We
define a type to be (Bishop-) finite if it is equal to some finite set, and we setup the predicate $\isFin$ to do so.

\todo{Cite Brent Yorgey's thesis?}

@@

Using $\BAut$, we can talk about types that have a specified cardinality, for example, the subuniverse of 2-element sets
is given by $\BAut[\Bool]$. This can be used to give the denotational semantics of a 1-bit reversible
language~\cite{caretteReversibleProgramsUnivalent2018}.

@@

\begin{corollary}
  $\UFin$ is a pointed, connected, 1-type, that is, a 1-groupoid. For every $n:\Nat$, \( \loopspace[\UFin[n],F_{n}] \eqv
  \Aut[\Fin[n]] \).
\end{corollary}
\note{Diagram below?}
% \[
%   \begin{tikzcd}
%     F_{0}
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
%     & F_{1}
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
%     & F_{2}
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=8em, in=125, out=55]
%     & \ldots
%     & F_{n}
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=8em, in=125, out=55]
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=12em, in=135, out=45]
%     & \ldots
%   \end{tikzcd}
% \]

>>>>

HoTT provides a rich internal language for working with weak higher groupoids, using the ``Types are weak
$\infty$-groupoids'' correspondence. Using the identity type for morphisms, a 1-groupoid in HoTT has points for 0-cells,
1-paths for 1-cells, and 2-paths for 2-cells, but there is at most one 2-cell between compatible 1-cells.~\footnote{This
is a locally-strict $(2,0)$-category, since every cell above 0 is invertible, and the hom-categories are posets, that
is, truth-value enriched.} The groupoid $\BFin$ can be written in HoTT as a type $\UFin$, which is a 1-groupoid. In this
section, we develop the tools necessary to construct $\UFin$, starting with a quick review of HoTT.

@@

The weak groupoid structure of types is given by the identity type in HoTT, where points are given by terms of the type,
and higher morphisms are given by the iterated identity type. Functions between types are functors between groupoids,
and type families, or functions to the universe, are indexed families of groupoids. A type family $P : A \to \UU$ comes
equipped with a functor $\pi_1 : \dsum{x:A}{P(x)} \to \UU$ which has a lifting operation giving it the structure of a
fibration (see the appendix for details).

\begin{toappendix}
\subsubsection{Identity Types}

@@

  \begin{definition}[Path Induction]
    \label{def:path-induction}
    Given a type family $C : \dfun{x,y:A}{(x \id_A y)} \to \UU$, and a function $c : \dfun{x:A}{C(x,x,\refl_x)}$, there is
    a function $f : \dfun{x,y:A}{\dfun{p:x \id_A y}{C(x,y,p)}}$ such that $f(x,x,\refl_x) \defeq c(x)$.
  \end{definition}

@@

The $\term{transport}$ operation lifts paths in the base space to maps between the fibers. Using the groupoid structure
of $A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name
this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.

@@

The type families (or fibrations) we're interested in are the ones where paths in the base space completely determine
the equivalences in the fibers -- these are called univalent fibrations.

@@

  \begin{definition}[Propositional Truncation]
    \label{def:prop-trunc}
    Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, or simply $\Trunc{A}$, is a higher inductive type
    generated by the following constructors,
    \begin{itemize}
      \item an inclusion function $\trunc{\blank} : A \to \Trunc{A}$,
      \item for each $x, y : \Trunc{A}$, a path $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$,
    \end{itemize}
    such that, given any type $B$ with
    \begin{itemize}
      \item a function $g : A \to B$,
      \item for each $x, y : B$, a path $\term{trunc*}(x,y) : x \id_{B} y$,
    \end{itemize}
    there is a unique function $f : \Trunc{A} \to B$ such that,
    \begin{itemize}
      \item $f(\trunc{a}) \equiv g(a)$
      \item for each $x, y : \Trunc{A}$, $\ap{f}{\term{trunc}(x,y)} \id_{B} \term{trunc*}(f(x),f(y))$.
    \end{itemize}
  \end{definition}

@@

  \begin{definition}[Set Quotient]
    \label{def:set-quot}
    Given a type $A$ which is an $\hSet$, and a relation $R : A \to A \to \hProp$, the set-quotient $\quot{A}{R}$ is the
    higher inductive type generated by
    \begin{itemize}
      \item an inclusion function $q : A \to \quot{A}{R}$,
      \item for each $x, y : A$ such that $R(x,y)$, a path $q(x) \id_{\quot{A}{R}} q(y)$,
      \item a set truncation, for each $x, y : \quot{A}{R}$ and $r, s : x \id_{\quot{A}{R}} y$, we have $r \id s$,
    \end{itemize}
    with \review{an appropriate induction principle.}
  \end{definition}

@@

Starting from a univalent universe which classifies all types, we want to define a subuniverse which classifies only
certain types, for example, types that satisfy some desired property. We use a prop-valued type family, that is, a
predicate on the universe, which picks out only those types, and collect them into a univalent subuniverse. Being
univalent ensures that the equality type of the ambient universe is reflected in the subuniverse.

@@

\begin{definition}[Subuniverse]
  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $P(X)$ is a proposition
  for every $X$. A subuniverse generated by $P$ is given by its total space $U \defeq \dsum{X:\UU}{P(X)}$ and $\El
  \defeq \fst$.
\end{definition}

\begin{proposition}[Univalent Subuniverse]
  Subuniverses generated by predicates are univalent.
\end{proposition}

@@

The types we're interested in are the finite types. In constructive mathematics, the notion of finiteness is subtle. We
define a type to be (Bishop) finite if it is equal to some finite set, and we setup the predicate $\isFin$ to do so.
\todo{Cite Brent Yorgey's thesis?}

@@

Using $\BAut$, we can talk about types that have a specified cardinality, for example, the subuniverse of 2-element sets
is given by $\BAut[\Bool]$. This has been used to construct the real projective spaces in
HoTT~\cite{buchholtzRealProjectiveSpaces2017}, and also to give the denotational semantics for a 1-bit reversible
language~\cite{caretteReversibleProgramsUnivalent2018}.

@@

\begin{corollary}
  $\UFin$ is a pointed, connected, 1-groupoid. For every $n:\Nat$, \( \loopspace[\UFin[n],F_{n}] \eqv \Aut[\Fin[n]] \).
\end{corollary}

The loopspace of a pointed type automatically has the structure of a group, that is, with $\refl_{\pt}$ for the neutral
element, path composition for the group multiplication, and path inverse for the group inverse operation. The group
axioms are then given by the higher paths corresponding to groupoid laws. We have shown that loops in $\UFin$ exactly
encode the automorphism group $\Aut[\Fin[n]]$ for every $n$. This is a general technique called delooping, where a group
can be identified with a 1-object groupoid, internally in HoTT.~\footnote{This technique also allows defining higher
groups in HoTT, as done in~\cite{buchholtzHigherGroupsHomotopy2018}.}

% \[
%   \begin{tikzcd}
%     F_{0}
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
%     & F_{1}
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
%     & F_{2}
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=8em, in=125, out=55]
%     & \ldots
%     & F_{n}
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=8em, in=125, out=55]
%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=12em, in=135, out=45]
%     & \ldots
%   \end{tikzcd}
% \]
